Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow

### 2025-02-26 manuscript/Manuscript.qmd
text

Central to ecosystem condition accounts are *variables* and *indicators*.
These are metrics chosen to reflect the central condition characteristics of the ecosystems.
These metrics are quantified and ideally monitored over time to reflect the status and trends in condition.
Indicators are variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e. that the variable equals the upper reference value which again reflect the value of the variables under the reference (pristine) condition.
Similarly, a value of 0 means that the variable is in the worst possible state, for example that the species or ecosystem function in question is completely lost from the ecosystem.
The reference condition needs to be defined for each ecosystem condition assessment separately, but SEEA EA gives some suggestions, such as an an ecosystem with no or minimal anthropogenic disturbance.

The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the ecosystem assets [@unitednations2021 §5.54].
Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies systematically along gradients of anthropogenic pressures and hence ecosystem condition.
The recommendation for spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to reliably estimate an indicator value for each ecosystem asset.
This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each ecosystem asset will have enough data to estimate its unique indicator value.
This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer* [@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
There are at least three ways to achieve this same complete areal coverage from indicator values that originally are spatially incomplete:

@@

>>>>

Central to ecosystem condition accounts are *variables* and *indicators*.
These are metrics chosen to reflect the central condition characteristics of the ecosystems.
These metrics are quantified and ideally monitored over time to reflect the status and trends in condition.
Indicators are variables that are normalised (referred to as rescaling in SEEA EA) against upper (X~100~) and lower (X~0~)reference levels to become bound between the values 0 and 1 [@czúcz].
This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e. that the variable equals the upper reference level, which again reflect the value of the variables under the reference condition.
Similarly, a value of 0 means that the variable is in the worst possible state, for example that the species or ecosystem function in question is completely lost from the ecosystem.
The reference condition needs to be defined for each ecosystem type and each ecosystem condition assessment separately, but SEEA EA gives some suggestions, such as an an ecosystem with no or minimal anthropogenic disturbance.

The SEEA EA recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded, or estimated, uniquely for each ecosystem asset so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the ecosystem assets [@unitednations2021 §5.54].
Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies systematically along gradients of anthropogenic pressures and hence ecosystem condition.
The recommendation for the spatial support, i.e. the spatial grain, also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to reliably estimate an indicator value for each ecosystem asset.
This problem does not exist for exhaustive, or complete, wall-to-wall data, like remotely sensed imagery, where each ecosystem asset will have enough data to estimate its unique indicator value.
This challenge if spatial generalisation <!--# use term upscaling and spatial aggregation -->is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer* [@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
There are at least three ways <!--# see UN guidelines annex D and Bierkens et al 2002 for many other ways --> to achieve this same complete areal coverage <!--# but we can get EAA estimated without inter-extrapolation. Need to frame this more using Bierkens et al --> from indicator values that originally are spatially incomplete:

@@

We acknowledge funding from the Research Council of Norway through the project EcoGaps (contract #320042).
